\documentclass{article}

\usepackage{../zwliew}

\title{MA2202 --- Algebra I}
\author{Liew Zhao Wei}
\date{Semester 2, 2022-2023}

\begin{document}
\maketitle
\hrule

\section{Preface}

This is a set of (incomplete) notes for a course on basic group theory.

\section{Groups}

We start by introducing groups.

\begin{definition}[Group]
  A \emph{group} is a set $G$ together with a law of composition that satisfies the following properties:
  \begin{itemize}
    \item \emph{Identity}: There is an element $e \in G$ such that for all $a \in G$, $ae = ea = a$.
    \item \emph{Associativity}: For all $a, b, c \in G$, $(ab)c = a(bc)$.
    \item \emph{Inverses}: For all $a \in G$, there exists $b \in G$ such that $ab = ba = e$.
  \end{itemize}
\end{definition}

\begin{definition}[Abelian group]
  A group $G$ is \emph{abelian} if its law of composition is commutative, that is, for all $a, b \in G$, $ab = ba$.
\end{definition}

\begin{remark}[Identity and inverse are unique]
  The identity and inverse of an element in a group are unique.
  We denote the identity by $e$, and the inverse of $a$ by $a^{-1}$.
\end{remark}

\begin{example}[Examples and non-examples of groups]
  Some examples of groups include:
  \begin{itemize}
    \item $\reals, \complex, \rationals, \integers$, and $\integers / n \integers$ are groups under addition.
    \item Removing the zero element, $\reals*, \complex*$, and $\rationals^*$ are groups under addition
    \item $\nats$ is not a group because inverses do not exist for any non-zero element.
  \end{itemize}
\end{example}

\begin{example}[General/special linear group]
  The \emph{general linear group} and \emph{special linear group} are groups of invertible matrices over a ring $R$ under matrix multiplication, denoted $\GL_n R$ and $\SL_n R$ respectively.
  $\GL_n R$ has non-zero determinant, while $\SL_n R$ has determinant 1.
\end{example}

\section{Symmetric Groups}

\subsection{Symmetric Groups}

\begin{definition}[Permutation]
  A \emph{permutation} of a set $S$ is a bijective map $p$ from $S$ to itself. That is, $p \colon S \to S$.
\end{definition}

\begin{definition}
  The set of all permutations of the indices $\{1, 2, \ldots, n\}$ is called the \emph{symmetric group} on $n$ elements, denoted by $S_n$.
\end{definition}

\begin{definition}
  Let $X$ be a set. The set of all permutations of $X$ is called the \emph{symmetric group} on $X$, denoted by $S_X$.
\end{definition}

\begin{proposition}
  $(S_n, \circ)$ and $(S_X, \circ)$ are isomorphic groups.
\end{proposition}

\subsection{Permutation Matrices}

A \emph{permutation matrix} is associated to any permutation.

\begin{definition}[Permutation matrix]
  An $n \times n$ permutation matrix is an $n \times n$ matrix $P$ whose columns are a permutation of the standard basis $e_1, e_2, \ldots, e_n$.
\end{definition}

\begin{proposition}
  Let $S_n^{''}$ denote the set of all $n \times n$ permutation matrices. Let $\times$ denote matrix multiplication. Then, $(S_n^{''}, \times)$ is a group and is isomorphic to $(S_n, \circ)$.
\end{proposition}

\subsection{Cyclic Notations}

We can express permutations as a product of disjoint cycles. This notation is called the \emph{cyclic notation} and is more compact compared to the tabular notation.

\begin{example}
  The permutation $p = (34)(12) \in S_4$ contains the following disjoint cycles: $1 \mapsto 2$, $2 \mapsto 1$, $3 \mapsto 4$, $4 \mapsto 3$, and $5 \mapsto 5$.
\end{example}

In fact, we can think of a cycle as a permutation.

\begin{definition}
  Consider a permutation $p \in S_n$ with a cycle ($x_1 x_2 \cdots x_r$). The cycle is a permutation $h \in S_n$ such that $h(x_1) = x_2, h(x_2) = x_3, \ldots, h(x_r) = h(x_1)$ and $h(y) = y$ for any $y$ not in the cycle.
\end{definition}

Then, a product of disjoint cycles simply refers to the composition of the corresponding permutations.

\begin{proposition}
  Every permutation can be factorized into a product of disjoint cycles. The factorization is unique up to an ordering of the product of cycles.
\end{proposition}

\subsection{Transpositions}

\begin{definition}
  A \emph{transposition} is a cycle of the form $h = (ij)$.
\end{definition}

\begin{proposition}
  Any cycle can be written as a product of transpositions.
\end{proposition}

\begin{proof}[Proof sketch]
  Let $h = (i_1, \ldots, i_r)$. Then, $(i_1 i_r)(i_1 i_{r-1})\cdots(i_1 i_2)$ is an equivalent permutation.
\end{proof}

Since any permutation can be expressed as a product of disjoint cycles, we can also express any permutation as a product of transpositions.

\begin{proposition}
  Every permutation can be expressed as a product of transpositions.
\end{proposition}

\subsection{The Sign Character}

\begin{lemma}
  For any permutation matrices $F$ and $H$ in $S_n^{''}$, we have $\det(FH) = \det(F)\det(H)$ and $\det(F) = \pm 1$.
\end{lemma}

Since permutations in $S_n$ are isomorphic to permutation matrices in $S_n^{''}$, observe the \emph{sign} of a permutation is equivalent to the determinant of its corresponding permutation matrix.

\begin{example}[Sign of a Permutation]
  Let $p$ be a permutation in $S_n$ and $\varphi$ be an isomorphism between $S_n$ and $S_n^{''}$. Then, the \emph{sign} of $p$ is $\mathsf{sgn}(p) = \det(\varphi(p))$.
\end{example}

However, the determinant of a matrix is typically defined in terms of its sign, so we define the sign in an alternative fashion.

\begin{definition}
  Let $\mathbf{x} = (x_1, x_2, \ldots, x_n)$ be $n$ variables.

  Define $P(\mathbf{x}) = P(x_1, \ldots, x_n) = \prod\limits_{1 \le i < j \le n}(x_i - x_j)$.

  Then, for each permutation $f \in S_n$, define $P_f(x_1, \ldots, x_n) = P(x_{f(1)}, x_{f(2)}, \ldots, x_{f(n)}) = \prod\limits_{1 \le i < j \le n}(x_{f(i)} - x_{f(j)})$.
\end{definition}

\begin{proposition}
  For any permutation $f \in S_n$, $P_f(x_1, \ldots, x_n) = \pm P(x_1, \ldots, x_n)$. We call the sign of $P_f$ the \emph{sign character} of $f$.
\end{proposition}

\begin{proposition}
  For any permutations $f, h \in S_n$, $\mathsf{sgn}(f \circ h) = \mathsf{sgn}(f) \mathsf{sgn}(h)$.
\end{proposition}

\begin{definition}
  A permutation $f \in S_n$ is \emph{even} if $\mathsf{sgn}(f) = 1$. Otherwise, it is \emph{odd}.
\end{definition}

\begin{theorem}
  Let $f \in S_n$ be a permutation. $f$ is even iff $f$ is a product of an even number of transpositions.
\end{theorem}

\begin{proposition}
  The set of all even permutations in $S_n$ is a subgroup of $S_n$.
\end{proposition}

\section{Cayley's Theorem}

\begin{theorem}[Cayley's Theorem]
  Every finite group of order $n$ is isomorphic to a subgroup of the symmetric group $S_n$.
\end{theorem}

Cayley's Theorem gives a way of realizing an abstract group as a subgroup of a more concrete group $S_n$. However, it is difficult to use in practice because the order of $S_n$ is usually too large in comparison to $n$. If $|G| = n$, then $|S_n| = n!$.

\section{Cosets}

\begin{definition}[Coset]
  Let $H$ be a subgroup of $G$ and $g \in G$ be an element of $G$. Then, a \textbf{left coset} is the subset
  $$
    gH = \{gh \mid h \in H\}
  $$
  A \textbf{right coset} is the subset
  $$
    Hg = \{hg \mid h \in H\}.
  $$
\end{definition}

Note that $eH = He = H$.



\end{document}
